# Introduction to mixed models

```{r}
#| label: setup
#| file: "_common.R"
#| include: true
#| message: false
options(digits=2)
```

This chapter considers tools for models with repeated measures from a modern perspective, using random effects for modelling. This class of model, called hierarchical models, multilevel models or mixed models in simple scenarios, give us more flexibility to account for complex scenarios in which there may be different sources of variability.

For example, consider a large-scale replication study about teaching methods. We may have multiple labs partaking in a research program and each has unique characteristics. Because of these, we can expect that measurements collected within a lab will be correlated. At the same time, we can have repeated mesures for participants in the study. One can view this setup as a hierarchy, with within-subject factor within subject within lab. In such settings, the old-school approach to analysis of variance becomes difficult, if not impossible; it doesn't easily account for the heterogeneity in the lab sample size and does not let us estimate the variability within labs.

We begin our journey with the same setup as for repeated measures ANOVA by considering one-way within-subject ANOVA model. We assign each participant (subject) in the study to all of the experimental treatments, in random order. If we have one experimental factor $A$ with $n_a$ levels, the model is
\begin{align*}\underset{\text{response}\vphantom{l}}{Y_{ij}} = \underset{\text{global mean}}{\mu_{\vphantom{j}}} + \underset{\text{mean difference}}{\alpha_j} + \underset{\text{random effect for subject}}{S_{i\vphantom{j}}} + \underset{\text{error}\vphantom{l}}{\varepsilon_{ij}}.
\end{align*}
In a random effect model, we assume that the subject effect $S_i$ is a random variable; we take $S_i \sim \mathsf{No}(0, \sigma^2_s)$ and the latter is assumed to be independent of the noise $\varepsilon_{ij} \sim \mathsf{No}(0, \sigma^2_e)$. The model parameters that we need to estimate are the global mean $\mu$, the mean differences $\alpha_1, \ldots, \alpha_{n_a}$, the subject-specific variability $\sigma^2_s$ and the residual variability $\sigma^2_e$, with the sum-to-zero constraint $\alpha_1 + \cdots + \alpha_{n_a}=0$.

Inclusion of random effects introduces positive correlation between measurements: specifically, the correlation between two observations from the same subject will be $\rho=\sigma^2_s/(\sigma^2_s+\sigma^2_e)$ and zero otherwise. This correlation structure is termed compound symmetry, since the correlation between measurements, $\rho$, is the same regardless of the order of the observations. If there are multiple random effects, the dependence structure will be more complicated.

In the repeated measure models, we need to first reduce measurements to a single average per within-subject factor, then fit the model by including the subject as a blocking factor. We are therefore considering subjects as fixed effects by including them as blocking factors, and estimate the mean effect for each subject: the value of $\sigma^2_s$ is estimated from the mean squared error of the subject term, but this empirical estimate can be negative. By contrast, the mixed model machinery will directly estimate the variance term, which will be constrained to be strictly positive.

## Fixed vs random effects


Mixed models include, by definition, both **random** and **fixed** effects. Fixed effects are model parameters corresponding to overall average or difference in means for the experimental conditions. These are the terms for which we want to perform hypothesis tests and compute contrasts. So far, we have only considered models with fixed effects. 

Random effects, on the other hand, assumes that the treatments are random samples from a population of interest. If we gathered another sample, we would be looking at a new set of treatments. Random effects model the variability arising from the sampling of that population and focuses on variance and correlation parameters. Addition of random effects does not impact the population mean, but induces variability and correlation within subject. There is no consensual definition, but @Gelman:2005 lists a handful:

> When a sample exhausts the population, the corresponding variable is fixed; when the sample is a small (i.e., negligible) part of the population the corresponding variable is random [Green and Tukey (1960)].

> Effects are fixed if they are interesting in themselves or random if there is interest in the underlying population (e.g., Searle, Casella and McCulloch [(1992), Section 1.4])

In terms of estimation, fixed effect terms are mean parameters, while all random effects will be obtained from variance and correlation parameters. In the repeated measure approach with fixed effects and blocking, we would estimate the average for each subject despite the fact that this quantity is of no interest. Estimating a mean with only a handful of measurements is a risky business and the estimated effects are sensitive to outliers. 

Random effects would proceed to directly estimate the variability arising from different subjects. We can still get predictions for the subject-specific effect, but this prediction will be shrunk toward the global mean for that particular treatment category. As we gather more data about the subjects, the predictions will become closer to the fixed effect estimates when the number of observations per subject or group increases, but these prediction can deviate from mean estimates in the case where there are few measurements per subject.


@Oehlert:2010 identifies the following step to perform a mixed model

1. Identify sources of variation
2. Identify whether factors are crossed or nested
3. Determine whether factors should be fixed or random
4. Figure out which interactions can exist and whether they can be fitted.

Sources of variations are all factors (including identifiers) which could influence the response. 

We say to factors are nested ($A$ within $B$) when one can only coexist within the levels of the other: this has implications, for we cannot have interaction between the two. In between-subject experiments, subjects are nested in between-subject factors and the experimental factors are crossed, meaning we can assign an experimental unit or a subject to each factor combination. Interactions can occur for the between-subject factors, although we need participants in each subcategory to estimate them.

:::{#exm-tempdist}

## Temporal distancing in peace prospects

We consider a three-way mixed design from Study 5 of @Halevy.Berson:2022. The research studied how temporal distance impacted the prospects of peace by presenting participants with a fictional scenario in which two tribes in a fictional country, Velvetia, where either at war or peace and asking what they think the outcome would be in the near and distance future. Participants rated the likeliness using a Likert scale  ranging from extremely unlikely (1) to extremely likely (7) for the question

>There is currently [war/peace] between the two tribes in Velvetia. Thinking about [next year/in 20 years], how likely is it that there will be [war/peace] in Velvetia?

Each participant was assigned to a single combination of the current state `curstate` and the predicted outcome `predout` and answered the question for both temporal horizon, `tempdist`. A rapid examination shows that we have a complete design: there are participants assigned to each subcategory of the between-subject factors.


```{r}
#| eval: true
#| echo: true
data(HB22_S5, package = "hecedsm")
xtabs(~ curstate + predout, data = HB22_S5)
```

1. The sources of variation are participants `id`, between-subject factors `predout` and `curstate` and within-subjects factor `tempdist`.
2. Between subject factors `predout` and `curstate` are crossed, whereas `tempdist` is nested within `id`.
3. `id` is random, whereas `predout`, `curstate` and `tempdist` are fixed effects
4. We can have interactions between the three experimental factors  `predout`, `curstate` and `tempdist` since the database contains instances of both of the former together, and since each person gets two `tempdist` levels. However, no interaction are possible between `id` and `tempdist` (due to lack of replication), or between between-subject factors and identifiers.

With data in long format, meaning each line contains a single response and characteristics are repeated per row, with an ` software will automatically capture the random effect. We can fit the equivalent of the repeated measure mixed ANOVA using a linear mixed model in **R**, by simply specifying a random intercept for the participant `id`. The full model with the three-way interaction includes eight components for the mean and two variant component (residual variance, and subject-specific variance).

```{r}
# In R, two packages offer linear mixed model fit via (RE)ML
# lme4 is more modern (and reliable?), but does not allow for unequal variant
lme4::lmer(likelihood ~ curstate*predout*tempdist + (1 | id),
     data = HB22_S5)
#  Note that in `lme4` package, the random effects are specified inside parenthesis
# nlme offers the possibility to acount for unequal variance per group
# Random effects are specified in `random` with formula notation, and a | to indicate
nlme::lme(likelihood ~ curstate*predout*tempdist, random = ~ 1 | id, data = HB22_S5)
```


:::



:::{#exm-happy-fakes-revisited}

## Happy fakes, remixed

We consider again the experiment of Amirabdolahian and Ali-Adeeb (2021) on smiling fakes and the emotion, this time from a pure mixed model perspective. This means we can simply keep all observations and model them accordingly. To fit the model, identifiers of subjects must be declared as factors (categorical variables). If we have repeated measurements of the within-subject factor, we may consider adding a random effect for the interaction between the subject and the within-subject factor.

```{r}
#| label: fig-aafulldat
#| fig-cap: "Jittered scatterplot of individual measurements per participant and stimulus type."
#| eval: true
#| out-width: '80%'
#| echo: false
ggplot(data = hecedsm::AA21,
       aes(x = id,
           group = stimulus,
           colour = stimulus,
           y = latency)) +
    geom_point(position = "jitter") +
  theme_classic() +
  theme(legend.position = "bottom")
```

@fig-aafulldat shows the raw measurements, including what are notable outliers that may be due to data acquisition problems or instrumental manipulations. Since the experiment was performed in a non-controlled setting (pandemic) with different apparatus and everyone acting as their own technician, it is unsurprising that the signal-to-noise ratio is quite small. We will exclude here (rather arbitrarily) measurements below a latency of minus 40.

```{r}
#| eval: true
#| echo: true
library(lmerTest) # Package to fit and tests for mixed models
options(contrasts = c("contr.sum", "contr.poly"))
mixedmod <- lmer(
  latency ~ stimulus + 
    (1 | id) + # random effect for subject
    (1 | id:stimulus), # interaction id and stimuluss
  # random effect for interaction 
  data = hecedsm::AA21 |> #remove outliers
    dplyr::filter(latency > -40))
# Output parameter estimates
print(mixedmod)
```

The model includes a fixed effect for stimulus, and variance terms for `stimulus` (nested within `id`) and subject identifier `id`.
We see that there is quite a bit of heterogeneity between participants and per stimulus participant pair, albeit less so for the interaction. All estimated variance terms are rather large. We can also look globally at the statistical evidence for the difference between the various stimuli.

```{r}
# Global effect of different faces
# ANOVA here is type III effects 
# computed from the 'lmerTest' package
anova(mixedmod)
```

The global $F$ test of significance for stimulus is based on an approximation that accounts for the correlation between observations; the denominator degrees of freedom for the approximate $F$ statistic are based on Satterthwaite's method. There is again no evidence of differences between experimental conditions. This is rather unsurprising if we look at the raw data in @fig-aafulldat.

:::

:::{#exm-verbalization-memorization}

## Verbalization and memorization

We consider a replication study of @Elliot:2021,
which studied verbalization and verbalization of kids aged 5, 6, 7 and 10. The replication was performed in 17 different school labs, adapting a protocol of @Flavell:1966, with an overall sample of 977 child partaking in the experiment.

Each participant was assigned to three tasks (`timing`): `delayed` recall with 15 seconds wait, or `immediate`, and finally  a naming task (`point-and-name`). The `taskorder` variable records the order in which these were presented: the order of `delayed` and `immediate` was counterbalanced across individuals, with the naming task always occurring last. The response variable is the number of words correctly recalled out of five. The experimenters also recorded the frequency at which students spontaneously verbalized during the task (except the naming task, where they were instructed to do so).

The timing is a within-subject factor, whereas task order and age are between-subject factors: we are particularly interested in the speech frequency and the improvement over time (pairwise differences and trend).

To fit the linear mixed model with a random effect for both children `id` and `lab`: since children are nested in lab, we must specify the random effects via `(1 | id:lab) + (1 | lab)` if `id` are not unique.

We modify the data to keep only 5 and 6 years old students, since most older kids verbalized during the task and we would have large disbalance (14 ten years old out of 235, and 19 out of 269 seven years old). We also exclude the point-and-name task, since verbalization was part of the instruction. This leaves us with 1419 observations and we can check that there are indeed enough children in each condition to get estimates.

```{r}
#| eval: true
#| echo: true
data(MULTI21_D2, package = "hecedsm")
MULTI21_D2_sub <- MULTI21_D2 |>
  dplyr::filter(
    age %in% c("5yo", "6yo"),
    timing != "point-and-name") |>
  dplyr::mutate(
    verbalization = factor(frequency != "never",
                           labels = c("no", "yes")),
    age = factor(age)) # drop unused age levels
xtabs(~ age + verbalization, data = MULTI21_D2_sub)
```

Given that we have multiple students of every age group, we can include two-way and three-way interactions in the $2^3$ design. We also include random effects for the student and the lab.

```{r}
#| eval: true
#| echo: true
library(lmerTest)
library(emmeans)
hmod <- lmer(
  mcorrect ~ age*timing*verbalization + (1 | id:lab) + (1 | lab), 
  data = MULTI21_D2_sub)
# Parameter estimates
#summary(hmod)
```

We focus here on selected part of the output from `summary()` giving the estimated variance terms.

```{r}
#| echo: true
#> Random effects:
#>  Groups   Name        Variance Std.Dev.
#>  id:lab   (Intercept) 0.3587   0.599   
#>  lab      (Intercept) 0.0625   0.250   
#>  Residual             0.6823   0.826   
#> Number of obs: 946, groups:  id:lab, 473; lab, 17
```


We can interpret the results as follows: the total variance is the sum of the  `id`, `lab` and `residual` variances components give us an all but negligible effect of lab with `r round(100*(hmod@theta[2]^2/(hmod@sigma^2 + sum(hmod@theta^2))),1)` percent of the total variance, versus `r round(100*(hmod@theta[1]^2/(hmod@sigma^2 + sum(hmod@theta^2))),1)` percent for the children-specific variability. Since there are only 17 labs, and most of the individual specific variability is at the children level, the random effect for lab doesn't add much to the correlation.

```{r}
#| eval: true
#| echo: true
anova(hmod, ddf = "Kenward-Roger")
# Check estimated marginal means for age
emm <- emmeans(hmod, specs = "age")
emm
# Pairwise differences
pairdiff <- emm |> pairs()
pairdiff
```


The type III ANOVA table shows that there is no evidence of interaction between task order, age and verbalization (no three-interaction) and a very small difference for timing and verbalization. Thus, we could compute the estimated marginal means (95% confidence interval) for age with an estimated correct number of words of `r summary(emm)$emmean[1]`  (`r summary(emm)$lower.CL[1]`, `r summary(emm)$upper.CL[1]`) words out of 5 for the 5 years olds and `r summary(emm)$emmean[2]` (`r summary(emm)$lower.CL[2]`, `r summary(emm)$upper.CL[2]`) words for six years old. Note that, despite the very large number children in the experiment, the degrees of freedom from the Kenward--Roger method are much fewer, respectively `r summary(emm)$df[1]` and `r summary(emm)$df[2]` for five and six years old.

The $t$-test for the pairwise difference of the marginal effect is `r -summary(pairdiff)$estimate` words with standard error `r summary(pairdiff)$SE`. Judging from the output, the degrees of freedom calculation for the pairwise $t$-test are erroneous --- they seem to be some average between the number of entries for the five years old (`r table(MULTI21_D2_sub$age)[1]`) and six years old (`r table(MULTI21_D2_sub$age)[2]`), but this fails to account for the fact that each kid is featured twice. Given the large magnitude of the ratio, this still amounts to strong result provided the standard error is correct.

We can easily see the limited interaction and strong main effects from the interaction plot in @fig-interactionrecall. The confidence intervals are of different width because of the sample inbalance.

```{r}
#| label: fig-interactionrecall
#| fig-cap: "Interaction plot for the recall task for younger children."
#| eval: true
#| echo: false
emm_full <- summary(emmeans(
  hmod, 
  specs = c("age", 
            "timing", 
            "verbalization")))
emm_full$age <- factor(rep(c(5L, 6L), 
                    length.out = 8))
emm_full <- tibble::as_tibble(emm_full)
library(ggplot2)
ggplot(data = emm_full,
       mapping = aes(x = age,
               color = verbalization,
               group = verbalization,
               y = emmean,
               ymax = emmean + 
                 qt(0.975, df = df)*SE,
               ymin = emmean - 
                 qt(0.975, df = df)*SE)
       ) + 
  geom_linerange(orientation = "x") +
  geom_point() +
  geom_line() + 
  facet_wrap(~ timing) +
  labs(x = "age group (years)", 
       y = "",
       title = "Mean number of correct words per task and age group",
       subtitle = "Older children and those who verbalize remember more.",
       caption = "Mean with 95% confidence intervals") + 
  theme_classic() + 
  theme(legend.position = "bottom")
```

:::
